---
title: "Stats 485 Paper 1 - Appendix (V2)"
author: "Eleanor King, elejking"
date: "2025-02-06"
output:
  pdf_document: default
---

## Stats 485 Paper 1 - Appendix

External Requirements: 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
mydata = read.csv("http://dept.stat.lsa.umich.edu/~bbh/s485/data/mobility2.csv")
```

Overview of goals of computations to follow: 

1. Create a function to calculate Agresti-Coull confidence intervals
2. Use the function to calculate a confidence interval for each commuting zone 
in order to determine the relative frequency of upward mobility.
3. Summarize and visualize the confidence intervals by commuting zone
4. Summarize and visualize the confidence intervals by region (midwest, northeast, south, west)

#### Confidence Interval Function (Agresti-Coull)

```{r}
ac_ci <- function(x, n, conf_level = 0.95) {
  z <- qnorm(1 - (1 - conf_level / 2))
  z_squared <- z^2
  x_tilde <- x + z_squared / 2
  n_tilde <- n + z_squared
  p_tilde <- x_tilde / n_tilde
  
  se <- sqrt(p_tilde * (1 - p_tilde) / n_tilde)
  
  lower <- p_tilde - z * se
  upper <- p_tilde + z * se
  
  return(c(lower = lower, upper = upper))
}
```

#### Calculating confidence intervals by commuting zone using the ac_ci function

```{r}
mydata <- mydata %>%
  mutate(
    p_hat = phat_upmover / n_lowstart,
    ci_lower = mapply(function(x, n) ac_ci(x, n)[1], phat_upmover, n_lowstart),
    ci_upper = mapply(function(x, n) ac_ci(x, n)[2], phat_upmover, n_lowstart)
  )
```

#### Summary of confidence intervals by zone

```{r}
summary_table <- mydata %>%
  select(zone, state, region, p_hat, ci_lower, ci_upper)

print(summary_table)
```
