---
title: "STATS 485 - Paper 2 Appendix"
author: "Eleanor King"
date: "2025-02-24"
output:
  pdf_document: default
  html_document: default
---

# GMP and Population Size

Overview of goals of computations to follow: 

1. Detail the statistical models and assumptions necessary for understanding and form hypotheses relating to the influence of variables other than population on per-capita GMP

2. Perform a preliminary exploration of the data set and introduce new variables necessary for evaluating hypotheses. 

3. Create a model to demonstrate the supra-linear relationship between GMP and population size. 

4. Introduce alternatives to the previously specified model that address the hypotheses formed at the outset of the appendix. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(caret)


msadata <- read_csv('http://dept.stat.lsa.umich.edu/~bbh/s485/data/gmp-2006-all.csv')

# Addressing missing data
msa_small = msadata |> dplyr::select(MSA:pop) |> na.omit()

# With missing data
msadata |> complete.cases() |> table()
# Without missing data
msa_small |> complete.cases() |> table()

# Addressing the error 'In pop * pcgmp : NAs produced by integer overflow'
#msa_small$pop <- as.double(msa_small$pop)
```


### Appendix I: Detail of Statistical Models


$$\text{If } Y \approx cN^b, \text{ for some } c > 0, b > 1, \text{then } \log(Y/N) 
\approx \beta_0 + \beta_1\log(N) \text{ for some } \beta_0 \in (-\infty, \infty),
\beta_1 > 0$$

$$\text{ Additionally, } \log(Y) \approx \beta_0 + (1 + \beta_1)\log(N)$$

##### Hypothesis I: Influence of the Finance Sector 
          
    Larger metropolitan areas tend to have a higher concentration of financial 
    firms. Because the careers in the financial sector are positively correlated 
    with wage, cities with a greater share of their economy in finance might 
    exhibit higher per capita GMP. Thus, if not properly taken into account, a 
    regression could falsely attribute this effect to population size 
    contributing to supra-linear scaling (Brookings). 

##### Hypothesis II: Influence of the ICT sector

    The information, communication, and technology (ICT) sector thrives in 
    metropolitan areas in part because larger cities attract highly skilled 
    professionals who are crucial for the ICT sector. If the ICT industry share 
    is positively correlated with per capita GMP, but not properly accounted 
    for in the regresion model, the coefficient on population size may absorb 
    some of this effect contributing to supra-linear scaling (Nature).

### Appendix II: Exploratory Analyses

```{r}
# Overall GMP
msa_small <- msadata |> 
  mutate(GMP = pcgmp * pop) |>
# Interaction term for finance and population
  mutate(finance_pop = finance * pop) |>
# Interaction term for ICT and population
  mutate(ict_pop = ict * pop)

# GMP vs. Population Size
ggplot(msa_small, aes(x = pop, y = GMP)) +
  geom_point() + 
  geom_smooth(formula = 'y ~ x', method = "loess", se = FALSE) +
  ggtitle("GMP vs. Population Size")

# Log(GMP) vs. Population Size
ggplot(msa_small, aes(x = pop, y = log(GMP))) +
  geom_point() + 
  geom_smooth(formula = 'y ~ x', method = "loess", se = FALSE) +
  ggtitle("Log(GMP) vs. Population Size")

# GMP vs. Log(Population Size)
ggplot(msa_small, aes(x = log(pop), y = GMP)) +
  geom_point() + 
  geom_smooth(formula = 'y ~ x', method = "loess", se = FALSE) +
  ggtitle("GMP vs. Log(Population Size)")

# Log(GMP) vs. Log(Population Size)
ggplot(msa_small, aes(x = log(pop), y = log(GMP))) +
  geom_point() + 
  geom_smooth(formula = 'y ~ x', method = "loess", se = FALSE) +
  ggtitle("Log(GMP) vs. Log(Population Size)")

```

The log GMP vs. log population size is the better scale for capturing patterns in the data using a regression model of relatively simple structure. 

```{r}
# By Finance Sector Strength
plot1 <- ggplot(msa_small, aes(x = log(pop), y = log(GMP), color = finance)) +
  geom_point(alpha = 0.8, size = 3) +
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "Log(GMP) vs. Log(Pop) - Finance Influence",
       x = "Log(pop)",
       y = "Log(GMP)",
       color = "Finance Share")

plot1

# By ICT Strength
plot2 <- ggplot(msa_small, aes(x = log(pop), y = log(GMP), color = ict)) +
  geom_point(alpha = 0.8, size = 3) +
  scale_color_gradient(low = "green", high = "purple") +
  labs(title = "Log(GMP) vs. Log(Pop) - ICT Influence",
       x = "Log(pop)",
       y = "Log(GMP)",
       color = "ICT Share")

plot2
```

### Appendix III: Fitting the Power Law Model

```{r}
model <- lm(log(GMP) ~ log(pop), data = msa_small)
summary(model)
```
The slope for log(pop) is greater than 1 suggesting supra-linear scaling

```{r}
# Regression plot
ggplot(msa_small, aes(x = log(pop), y = log(GMP))) +
  geom_point(alpha = 0.7, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Log(GMP) Regression on Log(Population",
       x = "Log(Population)",
       y = "Log(GMP)") + 
  theme_minimal()

# Residuals vs. Fitted Values
plot(model$fitted.values, residuals(model),
     xlab = "Fitted Values (Log(GMP))",
     ylab = "Residuals",
     main = "Residuals vs. Fitted Values") ; abline(h = 0)
```
The residual plot shows an amorphous blob indicating that the we can assume homoskedasticity.

```{r}
model_base10 <- lm(log10(GMP) ~ log10(pop), data = msa_small)

residuals <- residuals(model)
in_sample_loss <- sum(residuals^2)

in_sample_loss
```

### Appendix IV: Fitting and Assessment of Alternate Models

```{r}
# Overall GMP vs. Finance Sector Share and ICT Sector Share
model_finance <- lm(log(GMP) ~ finance + ict, data = msa_small)
summary(model_finance)

# Overall GMP vs. Finance Share and ICT Share Interactions with Population
model_interactions <- lm(log(GMP) ~ finance_pop + ict_pop, data = msa_small)
summary(model_interactions)
```

```{r}
# Proportions of Non-Missing Values
prop_non_missing <- msa_small |>
  summarize(
    prop_finance = mean(!is.na(finance)),
    prop_ict = mean(!is.na(ict)),
    prop_finance_pop = mean(!is.na(finance_pop)),
    prop_ict_pop = mean(!is.na(ict_pop))
  )

print(prop_non_missing)

# Proportion of Complete Observations
prop_complete <- mean(complete.cases(msa_small |> 
                                       select(pcgmp, pop, finance, ict, finance_pop, ict_pop)))

print(prop_complete)

# Data Set of Necessary Variables
msa_clean <- msa_small |> 
  select(GMP, pop, finance, ict, finance_pop, ict_pop) |>
  na.omit()

# Adding Log Transformed Variables Explicitly
msa_clean <- msa_clean |>
  mutate(log_GMP = log(GMP),
         log_pop = log(pop),
         log_finance_pop = log(finance_pop),
         log_ict_pop = log(ict_pop))

# Squared Error Loss Function
squared_error_loss <- function(model, data) {
  preds <- predict(model, newdata = data)
  loss <- sum((preds - log(data$GMP))^2)
  return(loss)
}

# Models Using Cleaned Data
model_1 <- lm(log_GMP ~ log_pop, data = msa_clean)
alt_model_1 <- lm(log_GMP ~ log_finance_pop + log_pop, data = msa_clean)
alt_model_2 <- lm(log_GMP ~ log_ict_pop + log_pop, data = msa_clean)
alt_model_3 <- lm(log_GMP ~ log_finance_pop + log_ict_pop + log_pop, data = msa_clean)

# Evaluating Performance
loss_model_1 <- squared_error_loss(model_1, msa_clean)
loss_alt_1 <- squared_error_loss(alt_model_1, msa_clean)
loss_alt_2 <- squared_error_loss(alt_model_2, msa_clean)
loss_alt_3 <- squared_error_loss(alt_model_3, msa_clean)

print(loss_model_1)
print(loss_alt_1)
print(loss_alt_2)
print(loss_alt_3)

```

### Appendix V: Additional calculations for paper version 2

```{r}
# Cross Validation
train_control <- trainControl(method = "cv", number = 10)

model_1 <- train(log_GMP ~ log_pop, data = msa_clean, method = "lm", trControl = train_control)
alt_model_1 <- train(log_GMP ~ log_finance_pop + log_pop, data = msa_clean, method = "lm", trControl = train_control)
alt_model_2 <- train(log_GMP ~ log_ict_pop + log_pop, data = msa_clean, method = "lm", trControl = train_control)
alt_model_3 <- train(log_GMP ~ log_finance_pop + log_ict_pop + log_pop, data = msa_clean, method = "lm", trControl = train_control)

errors <- data.frame(model_1$results$RMSE, alt_model_1$results$RMSE, alt_model_2$results$RMSE, alt_model_3$results$RMSE)

print(errors)
```